---
title: "tenaciousD data analysis"
author: "Ross Cunning"
date: "April 17, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
```

-----

## Load libraries and prepare data for analysis
The "setup.R" script loads required libraries and imports data. Total S/H ratios and proportion clade D are calculated for each sample, and corals are categorized as C-dominated or D-dominated based on their dominant symbiont at time zero. Relative changes in Fv/Fm are calculated for each sample. After sourcing this script, the data are ready to be analyzed.
```{r run_setup, include=FALSE}
source("setup.R")
```

### Experimental design
Number of C-dominated and D-dominated coral cores in each treatment
```{r all_samples}
n <- with(subset(data, time==0), table(dom, ramp))
n <- addmargins(n)
pander(n, style="grid")
```

-----

## Analyze Fv/Fm data
We utilize generalized additive mixed models (GAMMs) to model relative Fv/Fm values as a smooth function of time, with dominant symbiont clade as a fixed effect and mother colony and core as random effects. We fit separate models for the cooling treatment and the heating treatment. After fitting the models, we simulate (n=1000) from the posterior distribution of the parameters of the fitted models and generate 84% confidence intervals on fitted values for visual display purposes (non-overlapping 84% confidence intervals are approximately equal to a significant difference at p < 0.05). To formally conduct this test as well to determing whether Fv/Fm differs between C-dominated and D-dominated corals on each date, we also generate a 95% confidence interval on the difference between these groups from the simulated values, and test whether it includes zero.

### Cooling treatment
```{r cool_gamm}
# Fit GAMM for cooling treatment
gm.cool <- gamm(rfvfm ~ dom + s(time, by=dom, k=4), random=list(mother=~1, sample=~1), 
                data=subset(data, ramp=="cool"), na.action=na.omit)
summary(gm.cool$gam)
# Get fitted values
pdat.cool <- expand.grid(time=seq(0,63,1), ramp=factor("cool"), dom=factor(c("C", "D")))
gm.cool.pred <- data.frame(cbind(pdat.cool, fit=predict(gm.cool$gam, pdat.cool)))
# Simulate from multivariate normal distribution of fitted model
set.seed(789)
Rbeta <- mvrnorm(n = 1000, coef(gm.cool$gam), vcov(gm.cool$gam))
Xp <- predict(gm.cool$gam, newdata = pdat.cool, type = "lpmatrix")
sim <- Xp %*% t(Rbeta)
# Extract 84% confidence intervals from simulation
gm.cool.pred$lci <- apply(sim, 1, quantile, 0.08)
gm.cool.pred$uci <- apply(sim, 1, quantile, 0.92)
```

#### Plot raw data with fitted model
Here, Fv/Fm data (mean ± s.d.) are plotted for clade C- and D-dominated corals, along with the fitted values from the GAMM and 95% confidence intervals.

```{r cool_gamm_plot, echo=FALSE}
par(mar=c(6,3,1,1), mgp=c(2,0.5,0), tck=-0.06, xpd=NA)
plot(NA, xlim=c(0,63), ylim=c(0, 1), bty="n", tck=-0.03, xaxt="n", ann=F)
mtext(side=2, text="Relative Fv/Fm", cex=1, line=1.75)
with(subset(gm.cool.pred, dom=="C"), {
  addpoly(time, uci, lci, col=alpha("blue", 0.4))
  lines(time, fit)
})
with(subset(gm.cool.pred, dom=="D"), {
  addpoly(time, uci, lci, col=alpha("red", 0.4))
  lines(time, fit)
})
# Plot raw data +/- standard deviation
pl <- lapply(datlist[["cool"]], function(dom) {
  arrows(dom$time[-1], dom$rfvfm.mean[-1] + dom$rfvfm.sd[-1], dom$time[-1], dom$rfvfm.mean[-1] - dom$rfvfm.sd[-1], 
         code=3, angle=90, length=0.05, xpd=NA, col=list("C"="blue", "D"="red")[[dom$dom[1]]])
  points(dom$rfvfm.mean ~ dom$time, pch=21, bg=list("C"="blue", "D"="red")[[dom$dom[1]]], ylim=c(0, 1), cex=1)
})
legend("bottomleft", legend=c("Corals initially dominated by clade C", "Corals initially dominated by clade D"),
       pch=21, pt.bg=c("blue","red"), bty="n", inset=0.1)
# Add x-axis
axis(side=1, at=seq(0,63,7), labels=seq(0,63,7), line=2, tck=-0.03)
axis(side=1, at=seq(0,63,7), labels=NA, lwd=0, lwd.ticks=par("lwd"), line=2, tck=0.03)
mtext(side=1, text="days (below) and temperature (°C, above)", line=4)
axis(side=1, at=seq(0,62,7)+3.5, labels=paste0(seq(23,15,-1), "°"), tick=F, line=0.5)
```

#### Test if Fv/Fm is the same in C- and D-dominated corals at each sampling date
```{r cool_gamm_CvsD}
# Calculate 95% confidence interval on difference between C and D corals at each sampling dates from simulation
sim1 <- cbind(pdat.cool, sim)
sim1 <- sim1[sim1$time %in% c(0,7,14,21,28,35,42,49,56,63), ]
sim1split <- split(sim1, f=sim1$time, drop=T)
CDdiff.95CI <- lapply(sim1split, function(x) quantile(apply(x[,4:ncol(x)], 2, diff), c(0.025, 0.975)))
# Test if 95% CI on difference between C and D corals contains zero
cool.test0 <- lapply(CDdiff.95CI, FUN=function(x) (x[1] < 0 & 0 < x[2])[[1]])
t(matrix(cool.test0, dimnames=list(day=names(cool.test0), Ho="C-dom = D-dom")))
```
Relative Fv/Fm is significantly higher (p < 0.05) in C-dominated corals than D-dominated corals on day 7 onwards.

### Heating treatment
```{r heat_gamm}
# Fit GAMM for heating treatment
gm.heat <- gamm(rfvfm ~ dom + s(time, by=dom, k=4), random=list(mother=~1, sample=~1), 
                data=subset(data, ramp=="heat"), na.action=na.omit)
summary(gm.heat$gam)
# Get fitted values
pdat.heat <- expand.grid(time=seq(0,42,1), ramp=factor("heat"), dom=factor(c("C", "D")))
gm.heat.pred <- data.frame(cbind(pdat.heat, fit=predict(gm.heat$gam, pdat.heat)))
# Simulate from multivariate normal distribution of fitted model
set.seed(789)
Rbeta <- mvrnorm(n = 1000, coef(gm.heat$gam), vcov(gm.heat$gam))
Xp <- predict(gm.heat$gam, newdata = pdat.heat, type = "lpmatrix")
sim <- Xp %*% t(Rbeta)
# Extract 95% confidence intervals of simulated values for plotting
gm.heat.pred$lci <- apply(sim, 1, quantile, 0.08)
gm.heat.pred$uci <- apply(sim, 1, quantile, 0.92)
```

#### Plot raw data with fitted model
Here, Fv/Fm data (mean ± s.d.) are plotted for clade C- and D-dominated corals, along with the fitted values from the GAMM and 95% confidence intervals.

```{r heat_gamm_plot, echo=FALSE} 
par(mar=c(6,3,1,1), mgp=c(2,0.5,0), tck=-0.06, xpd=NA)
plot(NA, xlim=c(0,63), ylim=c(0, 1), bty="n", tck=-0.03, xaxt="n", ann=F)
mtext(side=2, text="Relative Fv/Fm", cex=1, line=1.75)
with(subset(gm.heat.pred, dom=="C"), {
  addpoly(time, uci, lci, col=alpha("blue", 0.4))
  lines(time, fit)
})
with(subset(gm.heat.pred, dom=="D"), {
  addpoly(time, uci, lci, col=alpha("red", 0.4))
  lines(time, fit)
})
# Plot raw data +/- standard deviation
pl <- lapply(datlist[["heat"]], function(dom) {
  arrows(dom$time[-1], dom$rfvfm.mean[-1] + dom$rfvfm.sd[-1], dom$time[-1], dom$rfvfm.mean[-1] - dom$rfvfm.sd[-1], 
         code=3, angle=90, length=0.05, xpd=NA, col=list("C"="blue", "D"="red")[[dom$dom[1]]])
  points(dom$rfvfm.mean ~ dom$time, pch=21, bg=list("C"="blue", "D"="red")[[dom$dom[1]]], ylim=c(0, 1), cex=1)
})
legend("bottomleft", legend=c("Corals initially dominated by clade C", "Corals initially dominated by clade D"),
       pch=21, pt.bg=c("blue","red"), bty="n", inset=0.1)
# Add x-axis
axis(side=1, at=seq(0,63,7), labels=seq(0,63,7), line=2, tck=-0.03)
axis(side=1, at=seq(0,63,7), labels=NA, lwd=0, lwd.ticks=par("lwd"), line=2, tck=0.03)
mtext(side=1, text="days (below) and temperature (°C, above)", line=4)
axis(side=1, at=seq(0,41,7)+3.5, labels=paste0(seq(30,35,1), "°"), tick=F, line=0.5)
```

#### Test if Fv/Fm is the same in C- and D-dominated corals at each sampling date

```{r heat_gamm_CvsD}
# Calculate 95% confidence interval on difference between C and D corals at select dates (=sampling dates)
sim1 <- cbind(pdat.heat, sim)
sim1 <- sim1[sim1$time %in% c(0,7,14,21,28,35,42), ]
sim1split <- split(sim1, f=sim1$time, drop=T)
CDdiff.95CI <- lapply(sim1split, function(x) quantile(apply(x[,4:ncol(x)], 2, diff), c(0.025, 0.975)))
# Test if 95% CI on difference between C and D corals contains zero
heat.test0 <- lapply(CDdiff.95CI, FUN=function(x) (x[1] < 0 & 0 < x[2])[[1]])
t(matrix(heat.test0, dimnames=list(day=names(heat.test0), Ho="C-dom = D-dom")))
```
Relative Fv/Fm is significantly higher (p < 0.05) in D-dominated corals than C-dominated corals at day 7, 14, and 21.

------

## Analyze symbiont abundance (S/H ratios)
Symbiont abundance is analyzed for each treatment using a linear mixed model with log-transformed total symbiont to host cell ratio (S/H ratio) as the response variable, with a fixed effect of time and dominant symbiont clade, and random effects of mother colony and core. After fitting the model, Dunnett's tests are performed within each group of corals to determine if the total S/H ratio is different from time zero. When significant differences were identified (p<0.05), the percent change in S/H ratio was calculated relative to time zero.

### Cooling treatment
```{r cool_SHmod}
# Fit linear mixed model
coolSHmod <- lmerTest::lmer(log10(tot.SH) ~ timef * dom + (1|mother/sample), data=subset(data, ramp=="cool"), na.action=na.omit)
summary(coolSHmod)
# Conduct Dunnett's post-hoc tests
cool.dunnett <- rbind(contrast(lsmeans::lsmeans(coolSHmod, specs=c("timef", "dom")), "dunnett", by="dom"))
# Calculate percent change relative to time zero and extract significant contrasts
cool.dunnett <- within(data.frame(summary(cool.dunnett)), {
  loss <- paste0(sprintf("%.1f", round(-(1-(10^estimate)) * 100, 1)), "%")
  day <- substr(contrast, 1, 2)
})
cool.diff <- cool.dunnett[cool.dunnett$p.value < 0.05, ]
cool.diff
```

#### Plot raw data and annotate significant changes in S/H ratio
Mean total S/H ratios ± s.d. are plotted for C-dominated and D-dominated corals. Significant differences within each group relative to time zero are indicated with asterisks and the percent change in S/H ratio.

```{r cool_SHmod_plot, echo=FALSE}
par(mar=c(5,3,0,1), xpd=NA)
plot(NA, xlim=c(0,63), ylim=c(-4, 0.5), bty="n", tck=-0.03, ylab="", xlab="", xaxt="n")
axis(side=1, at=seq(0,63,7), labels=seq(0,63,7), line=2, tck=-0.03)
axis(side=1, at=seq(0,63,7), labels=NA, lwd=0, lwd.ticks=par("lwd"), line=2, tck=0.03)
mtext(side=1, text="days (below) and temperature (°C, above)", line=4)
axis(side=1, at=seq(0,62,7)+3.5, labels=paste0(seq(23,15,-1), "°"), tick=F, line=-0.5)
mtext(side=2, text="log10 S/H ratio", line=2)
pl <- lapply(datlist[["cool"]], function(dom) {
  dom <- na.omit(dom)
  arrows(dom$time, dom$logtot.SH.mean + dom$logtot.SH.sd, dom$time, dom$logtot.SH.mean - dom$logtot.SH.sd, code=3, angle=90, length=0.05, xpd=NA,
         col=list("C"="blue", "D"="red")[[dom$dom[1]]])
  lines(dom$time, dom$logtot.SH.mean, type="l", col="black", lty=2, pch=21)
  points(dom$logtot.SH.mean ~ dom$time, pch=21, bg=list("C"="blue", "D"="red")[[dom$dom[1]]], ylim=c(0, 1), cex=1)
})
for (i in 1:nrow(cool.diff)) {
  with(cool.diff, {
    dom <- dom[i]
    points(as.numeric(day[i])+1, with(datlist[["cool"]][[dom]], logtot.SH.mean[time==day[i]]), pch="*", cex=2)
    text(day[i], with(datlist[["cool"]][[dom]], logtot.SH.mean[time==day[i]]), labels=loss[i], pos=c(2,2,2)[i])
  })
}
legend("bottomleft", legend=c("Corals initially dominated by clade C", "Corals initially dominated by clade D"),
       pch=21, pt.bg=c("blue","red"), bty="n", inset=0.1)
```

### Heating treatment
```{r heat_SHmod}
# Fit linear mixed model
heatSHmod <- lmerTest::lmer(log10(tot.SH) ~ timef * dom + (1|mother/sample), data=subset(data, ramp=="heat"), na.action=na.omit)
summary(heatSHmod)
# Conduct Dunnett's post-hoc tests
heat.dunnett <- rbind(contrast(lsmeans::lsmeans(heatSHmod, specs=c("timef", "dom")), "dunnett", by="dom"))
# Calculate percent change relative to time zero and extract significant contrasts
heat.dunnett <- within(data.frame(summary(heat.dunnett)), {
  loss <- paste0(sprintf("%.1f", round(-(1-(10^estimate)) * 100, 1)), "%")
  day <- substr(contrast, 1, 2)
})
heat.diff <- heat.dunnett[heat.dunnett$p.value < 0.05, ]
heat.diff
```

#### Plot raw data and annotate significant changes in S/H ratio
Mean total S/H ratios ± s.d. are plotted for C-dominated and D-dominated corals. Significant differences within each group relative to time zero are indicated with asterisks and the percent change in S/H ratio.

```{r heat_SHmod_plot, echo=FALSE}
par(mar=c(5,3,0,1), xpd=NA)
plot(NA, xlim=c(0,63), ylim=c(-4, 0.5), bty="n", tck=-0.03, ylab="", xlab="", xaxt="n")
axis(side=1, at=seq(0,63,7), labels=seq(0,63,7), line=2, tck=-0.03)
axis(side=1, at=seq(0,63,7), labels=NA, lwd=0, lwd.ticks=par("lwd"), line=2, tck=0.03)
mtext(side=1, text="days (below) and temperature (°C, above)", line=4)
axis(side=1, at=seq(0,41,7)+3.5, labels=paste0(seq(30,35,1), "°"), tick=F, line=-0.5)
mtext(side=2, text="log10 S/H ratio", line=2)
pl <- lapply(datlist[["heat"]], function(dom) {
  dom <- na.omit(dom)
  arrows(dom$time, dom$logtot.SH.mean + dom$logtot.SH.sd, dom$time, dom$logtot.SH.mean - dom$logtot.SH.sd, code=3, angle=90, length=0.05, xpd=NA,
         col=list("C"="blue", "D"="red")[[dom$dom[1]]])
  lines(dom$time, dom$logtot.SH.mean, type="l", col="black", lty=2, pch=21)
  points(dom$logtot.SH.mean ~ dom$time, pch=21, bg=list("C"="blue", "D"="red")[[dom$dom[1]]], ylim=c(0, 1), cex=1)
})
for (i in 1:nrow(heat.diff)) {
  with(heat.diff, {
    dom <- dom[i]
    points(as.numeric(day[i])+1, with(datlist[["heat"]][[dom]], logtot.SH.mean[time==day[i]]), pch="*", cex=2)
    text(day[i], with(datlist[["heat"]][[dom]], logtot.SH.mean[time==day[i]]), labels=loss[i], pos=c(2,2,2)[i])
  })
}
legend("topright", legend=c("Corals initially dominated by clade C", "Corals initially dominated by clade D"),
       pch=21, pt.bg=c("blue","red"), bty="n", inset=0)
```

------

## Analyze dynamics of dominant and non-dominant symbionts together
The previous analysis was on the total symbiont to host cell ratio, but did not distinguish between clade C and D symbionts. Clade D symbionts are present at lower levels in clade C-dominated corals, and vice versa. Here, we analyze the dynamics of both clades simultaneously within C- and D-dominated corals using a linear mixed model approach. To analyze the dynamics of both (dominant and non-dominant) clades within corals in each treatment, log-transformed clade-specific S/H ratios were modeled as a function of clade identity, time, and dominant clade identity, with mother colony and core as random effects. Significant changes in either clade C or D abundance within C- and D-dominated corals across consecutive sampling intervals were evaluated with the package lsmeans.

### Cooling treatment
```{r cool_CD}
# Rearrange data in order to use clade identity as fixed effect in linear model
df <- data[!is.na(data$tot.SH), ]
df2 <- melt(df, id.vars=c(1:6,9:14), value.name="SH")
colnames(df2)[colnames(df2)=="variable"] <- "clade"
# Fit linear mixed model for cooling treatment
coolmod <- lmerTest::lmer(log10(SH) ~ timef * dom * clade + (timef+dom+clade|mother/sample), data=subset(df2, ramp=="cool"))
print(summary(coolmod), correlation=FALSE)
# Test for differences across consecutive time points in both clades' S/H ratios
coolmod.lsm <- lsmeans(coolmod, specs=c("clade", "timef", "dom"))
coolmod.contr <- data.frame(summary(contrast(coolmod.lsm, "consec", by=c("dom", "clade"))))
coolmod.lsm <- data.frame(summary(coolmod.lsm))
# Identify significant differences and calculate percent change
coolmod.sigs <- coolmod.contr[coolmod.contr$p.value < 0.05, ]
coolmod.sigs$loss <- -(1-10^coolmod.sigs$estimate)
coolmod.sigs <- cbind(colsplit(coolmod.sigs$contrast, pattern="-", names=c("timef", "start")), coolmod.sigs)
coolmod.sigs
```

#### Plot clades C and D dynamics in cooling treatment
Fitted values for clade C and D S/H ratio (±SE) are plotted for C-dominated and D-dominated corals at each time point. Consecutive time points are connected by arrows, and significant differences between time points are indicated by the letter of the clade that was significantly different.

```{r cool_CD_plot, echo=FALSE, fig.height=6, fig.width=6}
par(mar=c(3,3,1,1), mgp=c(1.5,0.4,0), tcl=-0.3, xpd=F)
df <- dcast(coolmod.lsm, timef + dom ~ clade, value.var="lsmean")
df$C.se <- dcast(coolmod.lsm, timef + dom ~ clade, value.var="SE")$C.SH
df$D.se <- dcast(coolmod.lsm, timef + dom ~ clade, value.var="SE")$D.SH
df <- with(df, df[order(dom, timef), ])
plot(NA, xlim=c(-6,0), ylim=c(-6,0), xlab="Clade D (log10 S/H)", ylab="Clade C (log10 S/H)", xaxs="i", yaxs="i")
for (tot in 10^seq(-5,-1,1)) {
  D <- 10^seq(-6,-1,0.01)
  C <- tot-D
  lines(c(log10(D), rev(log10(C))), c(log10(C), rev(log10(D))), col="gray", lty=1)
}
abline(a=0,b=1,lty=2)
arrows(df$D.SH, df$C.SH+df$C.se, df$D.SH, df$C.SH-df$C.se, length=0.025, angle=90, code=3, col="black")
arrows(df$D.SH+df$D.se, df$C.SH, df$D.SH-df$D.se, df$C.SH, length=0.025, angle=90, code=3, col="black")
for (j in 2:nrow(df)) {
  if (as.numeric(df$timef)[j] > as.numeric(df$timef)[j-1]) {
    arrows(df$D.SH[j-1], df$C.SH[j-1], df$D.SH[j], df$C.SH[j], length=0.07, code=2, lwd=2, col=c("blue","red")[df$dom[j]])
  }
}

#df2 <- join(df, coolmod.sigs, c("timef", "dom"))
#which.sig <- which(rowSums(is.na(df2))==0)
#for (i in which.sig) {
#  text(x=(df2[i, "D.SH"] + df2[i-1, "D.SH"]) / 2, y=(df2[i, "C.SH"] + df2[i-1, "C.SH"]) / 2, #labels=bquote(bold(.(substr(df2[i, "clade"], 1, 1)))), cex=0.7)
#}

df2 <- join(df, coolmod.sigs, c("timef", "dom"))
df2 <- dcast(df2, timef + dom + C.SH + D.SH ~ clade, value.var="loss")
colnames(df2) <- c("timef", "dom", "C.SH", "D.SH", "Closs", "Dloss")
df2 <- with(df2, df2[order(dom, timef), ])
df2$sig <- ifelse(is.na(df2$Closs), ifelse(is.na(df2$Dloss), "", "D"), ifelse(is.na(df2$Dloss), "C", "CD"))
which.sig <- which(df2$sig!="")
for (i in which.sig) {
  text(x=(df2[i, "D.SH"] + df2[i-1, "D.SH"]) / 2, y=(df2[i, "C.SH"] + df2[i-1, "C.SH"]) / 2, labels=bquote(bold(.(df2[i, "sig"]))), cex=0.7)
}

xx <- seq(-5,-4,len=4)
arrows(x0=xx[-4], y0=-0.2, x1=xx[-1], code=2, length=0.07, lwd=1.5, col="blue")
arrows(x0=xx[-4], y0=-0.4, x1=xx[-1], code=2, length=0.07, lwd=1.5, col="red")
text(xx, -0.6, labels=c(24,20,18,15), cex=0.7, adj=c(0.5, 0.35))
text(-5.5, c(-0.2, -0.4, -0.6), labels=c("C-dom.", "D-dom.", "°C"), cex=0.75, adj=c(0.5, 0.35))
rect(-6,-0.75,-3.8,0)
```

### Heating treatment
```{r heat_CD}
# Rearrange data in order to use clade identity as fixed effect in linear model
df <- data[!is.na(data$tot.SH), ]
df2 <- melt(df, id.vars=c(1:6,9:14), value.name="SH")
colnames(df2)[colnames(df2)=="variable"] <- "clade"
# Fit linear mixed model for heating treatment
heatmod <- lmerTest::lmer(log10(SH) ~ timef * dom * clade + (timef+dom+clade|mother/sample), data=subset(df2, ramp=="heat"))
print(summary(heatmod), correlation=FALSE)
# Test for differences across consecutive time points in both clades' S/H ratios
heatmod.lsm <- lsmeans(heatmod, specs=c("clade", "timef", "dom"))
heatmod.contr <- data.frame(summary(contrast(heatmod.lsm, "consec", by=c("dom", "clade"))))
heatmod.lsm <- data.frame(summary(heatmod.lsm))
# Identify significant differences and calculate percent change
heatmod.sigs <- heatmod.contr[heatmod.contr$p.value < 0.05, ]
heatmod.sigs$loss <- -(1-10^heatmod.sigs$estimate)
heatmod.sigs <- cbind(colsplit(heatmod.sigs$contrast, pattern="-", names=c("timef", "start")), heatmod.sigs)
heatmod.sigs
```

#### Plot clades C and D dynamics in heating treatment
Fitted values for clade C and D S/H ratio (±SE) are plotted for C-dominated and D-dominated corals at each time point. Consecutive time points are connected by arrows, and significant differences between time points are indicated by the letter of the clade that was significantly different.

```{r heat_CD_plot, echo=FALSE, fig.height=6, fig.width=6}
par(mar=c(3,3,1,1), mgp=c(1.5,0.4,0), tcl=-0.3, xpd=F)
df <- dcast(heatmod.lsm, timef + dom ~ clade, value.var="lsmean")
df$C.se <- dcast(heatmod.lsm, timef + dom ~ clade, value.var="SE")$C.SH
df$D.se <- dcast(heatmod.lsm, timef + dom ~ clade, value.var="SE")$D.SH
df <- with(df, df[order(dom, timef), ])
plot(NA, xlim=c(-6,0), ylim=c(-6,0), xlab="Clade D (log10 S/H)",ylab="Clade C (log10 S/H)", xaxs="i", yaxs="i")
for (tot in 10^seq(-5,-1,1)) {
  D <- 10^seq(-6,-1,0.01)
  C <- tot-D
  lines(c(log10(D), rev(log10(C))), c(log10(C), rev(log10(D))), col="gray", lty=1)
}
abline(a=0,b=1,lty=2, col="gray")
arrows(df$D.SH, df$C.SH+df$C.se, df$D.SH, df$C.SH-df$C.se, length=0.025, angle=90, code=3, col="black")
arrows(df$D.SH+df$D.se, df$C.SH, df$D.SH-df$D.se, df$C.SH, length=0.025, angle=90, code=3, col="black")
for (j in 2:nrow(df)) {
  if (as.numeric(df$time)[j] > as.numeric(df$time)[j-1]) {
    arrows(df$D.SH[j-1], df$C.SH[j-1], df$D.SH[j], df$C.SH[j], length=0.07, code=2, lwd=2,
           col=c("blue","red","green")[df$dom[j]])
  }
}

df2 <- join(df, heatmod.sigs, c("timef", "dom"))
df2 <- dcast(df2, timef + dom + C.SH + D.SH ~ clade, value.var="loss")
colnames(df2) <- c("timef", "dom", "C.SH", "D.SH", "Closs", "Dloss")
df2 <- with(df2, df2[order(dom, timef), ])
df2$sig <- ifelse(is.na(df2$Closs), ifelse(is.na(df2$Dloss), "", "D"), ifelse(is.na(df2$Dloss), "C", "CD"))
which.sig <- which(df2$sig!="")
for (i in which.sig) {
  text(x=(df2[i, "D.SH"] + df2[i-1, "D.SH"]) / 2, y=(df2[i, "C.SH"] + df2[i-1, "C.SH"]) / 2, labels=bquote(bold(.(df2[i, "sig"]))), cex=0.7)
}

xx <- seq(-5,-4,len=3)
arrows(x0=xx[-3], y0=-0.2, x1=xx[-1], code=2, length=0.07, lwd=1.5, col="blue")
arrows(x0=xx[-3], y0=-0.4, x1=xx[-1], code=2, length=0.07, lwd=1.5, col="red")
text(xx, -0.6, labels=c(29,33,35), cex=0.7, adj=c(0.5, 0.35))
text(-5.5, c(-0.2, -0.4, -0.6), labels=c("C-dom.", "D-dom.", "°C"), cex=0.75, adj=c(0.5, 0.35))
rect(-6,-0.75,-3.8,0)
```

# Final figures
## Figure 1
The impact of incremental cooling (A.) and heating (B.) on relative Fv/Fm (upper panels) and total symbiont abundance (lower panels). For Fv/Fm, mean ± SD at each time point are plotted along with the fitted GAMM (solid lines) and 84% confidence intervals (shaded region) for initially C-dominated corals (blue) and D-dominated corals (red). Non-overlapping 84% confidence intervals indicate a significant difference between C and D corals at p < 0.05. For symbiont abundance, the mean of log-transformed total S/H ratios ± SD at each time point are plotted for intially C-dominated (blue) and D-dominated (red) corals. Significant differences within each group relative to initial values are indicated by an asterisk, along with the percent reduction in total S/H ratio. The horizontal axis displays days since the experiment began (below axis), and the temperature (°C) corals experienced during each weeklong increment (above axis).
```{r Fig1, echo=FALSE, fig.height=6.85039, fig.width=6.85039}
#pdf(file = "output/Figure1.pdf", width=6.85, height=6.85)
layout(mat=matrix(c(1,1,2,2,3,3,4,4), ncol=2))
# Cooling Fv/Fm
par(mar=c(0,3,3.5,0), mgp=c(1.5,0.25,0), tck=0.025, xpd=NA)
plot(NA, xlim=c(0,63), ylim=c(0, 1), bty="n", xaxt="n", yaxt="n", ann=F)
axis(side=2, at=seq(0,1,0.2), labels=c("0","0.2","0.4","0.6","0.8","1"), tck=0.025)
title("A. Cooling", adj=0, cex.main=1.5, line=1.5)
mtext(side=2, text="Relative Fv/Fm", cex=0.75, line=1.5)
with(subset(gm.cool.pred, dom=="C"), {
  addpoly(time, uci, lci, col=alpha("blue", 0.4))
  lines(time, fit)
})
with(subset(gm.cool.pred, dom=="D"), {
  addpoly(time, uci, lci, col=alpha("red", 0.4))
  lines(time, fit)
})

# Plot raw data +/- standard deviation
pl <- lapply(datlist[["cool"]], function(dom) {
  arrows(dom$time[-1], dom$rfvfm.mean[-1] + dom$rfvfm.sd[-1], dom$time[-1], dom$rfvfm.mean[-1] - dom$rfvfm.sd[-1], 
         code=3, angle=90, length=0.05, xpd=NA, col=list("C"="blue", "D"="red")[[dom$dom[1]]])
  points(dom$rfvfm.mean ~ dom$time, pch=21, bg=list("C"="blue", "D"="red")[[dom$dom[1]]], ylim=c(0, 1), cex=1)
})

#cool.sigdays <- as.numeric(names(cool.test0[cool.test0==F]))
#points(cool.sigdays, c(1.07,1.05,1.03,1.02,1,0.97,0.85,0.75,0.65), pch="*", cex=1.5)

# Cooling total S/H ratio
par(mar=c(3.5,3,0,0))
plot(NA, xlim=c(0,63), ylim=c(-4, 0.5), bty="n", tck=0.025, ylab="", xlab="", xaxt="n")
axis(side=1, at=seq(0,63,7), labels=seq(0,63,7), line=2, tck=0.025)
axis(side=1, at=seq(0,62,7)+3.5, labels=paste0(seq(23,15,-1), "°"), tick=F, line=0.6)
mtext(side=2, text="log10 S/H ratio", line=1.5, cex=0.75)
pl <- lapply(datlist[["cool"]], function(dom) {
  dom <- na.omit(dom)
  arrows(dom$time, dom$logtot.SH.mean + dom$logtot.SH.sd, dom$time, dom$logtot.SH.mean - dom$logtot.SH.sd, code=3, angle=90, length=0.05, xpd=NA,
         col=list("C"="blue", "D"="red")[[dom$dom[1]]])
  lines(dom$time, dom$logtot.SH.mean, type="l", col="black", lty=2, pch=21)
  points(dom$logtot.SH.mean ~ dom$time, pch=21, bg=list("C"="blue", "D"="red")[[dom$dom[1]]], ylim=c(0, 1), cex=1)
})
for (i in 1:nrow(cool.diff)) {
  with(cool.diff, {
    dom <- dom[i]
    points(as.numeric(day[i])+2, with(datlist[["cool"]][[dom]], logtot.SH.mean[time==day[i]]), pch="*", cex=1.5)
    text(day[i], with(datlist[["cool"]][[dom]], logtot.SH.mean[time==day[i]]), labels=loss[i], pos=c(2,2,2)[i])
  })
}

axis(side=1, at=-7, labels="°C", tick=F, line=0.6)
axis(side=1, at=-7, labels="days", tick=F, line=2)


# Heating Fv/Fm
par(mar=c(0,3,3.5,0))
plot(NA, xlim=c(0,63), ylim=c(0, 1), bty="n", tck=0.025, xaxt="n", yaxt="n", ann=F)
axis(side=2, at=seq(0,1,0.2), labels=c("0","0.2","0.4","0.6","0.8","1"), tck=0.025)
title("B. Heating", adj=0, cex.main=1.5, line=1.5)
with(subset(gm.heat.pred, dom=="C"), {
  addpoly(time, uci, lci, col=alpha("blue", 0.4))
  lines(time, fit)
})
with(subset(gm.heat.pred, dom=="D"), {
  addpoly(time, uci, lci, col=alpha("red", 0.4))
  lines(time, fit)
})
# Plot raw data +/- standard deviation
pl <- lapply(datlist[["heat"]], function(dom) {
  arrows(dom$time[-1], dom$rfvfm.mean[-1] + dom$rfvfm.sd[-1], dom$time[-1], dom$rfvfm.mean[-1] - dom$rfvfm.sd[-1], 
         code=3, angle=90, length=0.05, xpd=NA, col=list("C"="blue", "D"="red")[[dom$dom[1]]])
  points(dom$rfvfm.mean ~ dom$time, pch=21, bg=list("C"="blue", "D"="red")[[dom$dom[1]]], ylim=c(0, 1), cex=1)
})

# Heating total S/H ratio
par(mar=c(3.5,3,0,0), xpd=NA)
plot(NA, xlim=c(0,63), ylim=c(-4, 0.5), bty="n", tck=0.025, ylab="", xlab="", xaxt="n")
axis(side=1, at=seq(0,63,7), labels=seq(0,63,7), line=2, tck=0.025)
axis(side=1, at=seq(0,41,7)+3.5, labels=paste0(seq(30,35,1), "°"), tick=F, line=0.6)
pl <- lapply(datlist[["heat"]], function(dom) {
  dom <- na.omit(dom)
  arrows(dom$time, dom$logtot.SH.mean + dom$logtot.SH.sd, dom$time, dom$logtot.SH.mean - dom$logtot.SH.sd, code=3, angle=90, length=0.05, xpd=NA,
         col=list("C"="blue", "D"="red")[[dom$dom[1]]])
  lines(dom$time, dom$logtot.SH.mean, type="l", col="black", lty=2, pch=21)
  points(dom$logtot.SH.mean ~ dom$time, pch=21, bg=list("C"="blue", "D"="red")[[dom$dom[1]]], ylim=c(0, 1), cex=1)
})
for (i in 1:nrow(heat.diff)) {
  with(heat.diff, {
    dom <- dom[i]
    points(as.numeric(day[i])+2, with(datlist[["heat"]][[dom]], logtot.SH.mean[time==day[i]]), pch="*", cex=1.5)
    text(day[i], with(datlist[["heat"]][[dom]], logtot.SH.mean[time==day[i]]), labels=loss[i], pos=c(2,2,2)[i])
  })
}


```

## Figure 2
Dynamics of both dominant and subdominant symbionts in corals during incremental cooling (A.) and heating (B.) Fitted values of mean abundance of clade C and D symbionts in both C-dominated (blue) and D-dominated (red) corals from linear mixed models are plotted ± SE. Arrows connect consecutive time points corresponding to temperatures indicated in the legend. Letters at the midpoint of an arrow represent a statistically significant change in the abundance of the clade indicated at the specific time interval (p < 0.05). The dashed diagonal line represents equal abundances of clades C and D, and gray contours represent equal sum total abundance of C and D (i.e., constant total S/H) at order of magnitude intervals.
```{r Fig2, echo=FALSE, fig.height=3.425195, fig.width=6.85039}
par(mfrow=c(1,2), mar=c(2,2,0.5,0.5), mgp=c(1,0,0), tck=0.025)
df <- dcast(coolmod.lsm, timef + dom ~ clade, value.var="lsmean")
df$C.se <- dcast(coolmod.lsm, timef + dom ~ clade, value.var="SE")$C.SH
df$D.se <- dcast(coolmod.lsm, timef + dom ~ clade, value.var="SE")$D.SH
df <- with(df, df[order(dom, timef), ])
plot(NA, xlim=c(-6,0), ylim=c(-6,0), xlab="Clade D (log10 S/H)", ylab="Clade C (log10 S/H)", cex.lab=0.75, cex.axis=0.75, xaxs="i", yaxs="i")
for (tot in 10^seq(-5,-1,1)) {
  D <- 10^seq(-6,-1,0.01)
  C <- tot-D
  lines(c(log10(D), rev(log10(C))), c(log10(C), rev(log10(D))), col="gray", lty=1)
}
#title("A. Cooling", adj=0, cex.main=1, line=0.5)
text(par("usr")[1], par("usr")[4], expression(bold("A.  ")), xpd=NA, pos=2, outer=T)
abline(a=0,b=1,lty=2)
arrows(df$D.SH, df$C.SH+df$C.se, df$D.SH, df$C.SH-df$C.se, length=0.025, angle=90, code=3, col="black")
arrows(df$D.SH+df$D.se, df$C.SH, df$D.SH-df$D.se, df$C.SH, length=0.025, angle=90, code=3, col="black")
for (j in 2:nrow(df)) {
  if (as.numeric(df$timef)[j] > as.numeric(df$timef)[j-1]) {
    arrows(df$D.SH[j-1], df$C.SH[j-1], df$D.SH[j], df$C.SH[j], length=0.07, code=2, lwd=2, col=c("blue","red")[df$dom[j]])
  }
}

df2 <- join(df, coolmod.sigs, c("timef", "dom"))
df2 <- dcast(df2, timef + dom + C.SH + D.SH ~ clade, value.var="loss")
colnames(df2) <- c("timef", "dom", "C.SH", "D.SH", "Closs", "Dloss")
df2 <- with(df2, df2[order(dom, timef), ])
df2$sig <- ifelse(is.na(df2$Closs), ifelse(is.na(df2$Dloss), "", "D"), ifelse(is.na(df2$Dloss), "C", "CD"))
which.sig <- which(df2$sig!="")
for (i in which.sig) {
  text(x=(df2[i, "D.SH"] + df2[i-1, "D.SH"]) / 2, y=(df2[i, "C.SH"] + df2[i-1, "C.SH"]) / 2, labels=bquote(bold(.(df2[i, "sig"]))), cex=0.7)
}

xx <- seq(-5,-4,len=4)
arrows(x0=xx[-4], y0=-0.2, x1=xx[-1], code=2, length=0.07, lwd=1.5, col="blue")
arrows(x0=xx[-4], y0=-0.4, x1=xx[-1], code=2, length=0.07, lwd=1.5, col="red")
text(xx, -0.6, labels=c(24,20,18,15), cex=0.5, adj=c(0.5, 0.35))
text(-5.5, c(-0.2, -0.4, -0.6), labels=c("C-dom.", "D-dom.", "°C"), cex=0.6, adj=c(0.5, 0.35))
rect(-6,-0.75,-3.8,0)

# Heating
par(mar=c(2,2,0.5,0.5))
df <- dcast(heatmod.lsm, timef + dom ~ clade, value.var="lsmean")
df$C.se <- dcast(heatmod.lsm, timef + dom ~ clade, value.var="SE")$C.SH
df$D.se <- dcast(heatmod.lsm, timef + dom ~ clade, value.var="SE")$D.SH
df <- with(df, df[order(dom, timef), ])
plot(NA, xlim=c(-6,0), ylim=c(-6,0), xlab="Clade D (log10 S/H)",ylab="Clade C (log10 S/H)", cex.lab=0.75, cex.axis=0.75, xaxs="i", yaxs="i")
for (tot in 10^seq(-5,-1,1)) {
  D <- 10^seq(-6,-1,0.01)
  C <- tot-D
  lines(c(log10(D), rev(log10(C))), c(log10(C), rev(log10(D))), col="gray", lty=1)
}
#title("B. Heating", adj=0, cex.main=1, line=0.5)
text(par("usr")[1], par("usr")[4], expression(bold("B.  ")), xpd=NA, pos=2, outer=T)
abline(a=0,b=1,lty=2)
arrows(df$D.SH, df$C.SH+df$C.se, df$D.SH, df$C.SH-df$C.se, length=0.025, angle=90, code=3, col="black")
arrows(df$D.SH+df$D.se, df$C.SH, df$D.SH-df$D.se, df$C.SH, length=0.025, angle=90, code=3, col="black")
for (j in 2:nrow(df)) {
  if (as.numeric(df$time)[j] > as.numeric(df$time)[j-1]) {
    arrows(df$D.SH[j-1], df$C.SH[j-1], df$D.SH[j], df$C.SH[j], length=0.07, code=2, lwd=2,
           col=c("blue","red","green")[df$dom[j]])
  }
}

df2 <- join(df, heatmod.sigs, c("timef", "dom"))
df2 <- dcast(df2, timef + dom + C.SH + D.SH ~ clade, value.var="loss")
colnames(df2) <- c("timef", "dom", "C.SH", "D.SH", "Closs", "Dloss")
df2 <- with(df2, df2[order(dom, timef), ])
df2$sig <- ifelse(is.na(df2$Closs), ifelse(is.na(df2$Dloss), "", "D"), ifelse(is.na(df2$Dloss), "C", "CD"))
which.sig <- which(df2$sig!="")
for (i in which.sig) {
  text(x=(df2[i, "D.SH"] + df2[i-1, "D.SH"]) / 2, y=(df2[i, "C.SH"] + df2[i-1, "C.SH"]) / 2, labels=bquote(bold(.(df2[i, "sig"]))), cex=0.7)
}

xx <- seq(-5,-4,len=3)
arrows(x0=xx[-3], y0=-0.2, x1=xx[-1], code=2, length=0.07, lwd=1.5, col="blue")
arrows(x0=xx[-3], y0=-0.4, x1=xx[-1], code=2, length=0.07, lwd=1.5, col="red")
text(xx, -0.6, labels=c(29,33,35), cex=0.5, adj=c(0.5, 0.35))
text(-5.5, c(-0.2, -0.4, -0.6), labels=c("C-dom.", "D-dom.", "°C"), cex=0.6, adj=c(0.5, 0.35))
rect(-6,-0.75,-3.8,0)
```
